Effects of bariatric surgery and weight loss on resting state functional connectivity

### 2020-12-21 manuscript/draft_rsfmri.Rmd
include glass brain of networks and LLMMs into draft

### Aggregated FC values
To extract aggregated FC values, we first calculated the mean DMN and reward networks over all participants and time points, adjusted for age and sex. We used the modified SwE option and used global signal regressed data as input. Network masks were formed from all voxels with a clusterwise $p_{FWE} < 0.05$ and we extracted the average connectivity from these masks for all data points.

@@

Further, we evaluated the impact of our denoising pipelines. Table \@ref(tab:tableQcfc) shows that in the minimally preprocessed data, mFD and connectivity was significantly positively correlated (mean Spearman's $r = $`r round(res[1,1],2)`), with a strong distance dependence (mean Spearman's $r = $`r round(res[1,5],2)`).  Denoising procedures (AROMA, AROMA+CC, AROMA+CC+GSR) gradually reduced the correlation of mFD and connectivity, and also alleviated the distance dependency. AROMA+CC+GSR performed best, with almost no mFD-FC correlation (mean Spearman's $r = $`r round(res[4,1],2)`), and minimal positive distance dependence (mean Spearman's $r = $`r round(res[4,5],2)`). Yet, AROMA+CC also performed well, and we thus calculated our analysis models for both preprocessing schemes.

## Interaction effect of intervention and time (preregistered analysis)

### DMN/reward network with whole brain mask (preregistered analysis)

@@

@@

## Interaction effect group by time on aggregated FC (preregistered analysis)

### DMN for two timepoints
```{r calc DMN two tp, eval = FALSE, echo = FALSE}
R1 = lme4::lmer(DMN_FC ~ timepoint*group + (1|subj))
R0 = lme4::lmer(DMN_FC ~ timepoint + group + (1|subj))
```
### DMN for two timepoints, adjusted for age, sex, mean FD
```{r calc DMN two tp adjusted, eval = FALSE, echo = FALSE}
R1 = lme4::lmer(DMN_FC~ timepoint*group + age + sex + meanFD (1|subj))
R0 = lme4::lmer(DMN_FC ~ timepoint + group + age + sex + meanFD (1|subj))
```
### DMN for three timepoints, adjusted for age, sex, mean FD
```{r calc DMN three tp, eval = FALSE, echo = FALSE}
R1 = lme4::lmer(DMN_FC~ timepoint*group + age + sex + meanFD (1|subj))
R0 = lme4::lmer(DMN_FC ~ timepoint + group + age + sex + meanFD (1|subj))
```

### reward
```{r calc rew two tp, eval = FALSE, echo = FALSE}
R1 = lme4::lmer(rew_FC ~ timepoint*group + (1|subj))
R0 = lme4::lmer(rew_FC ~ timepoint + group + (1|subj))
```
### reward for two timepoints, adjusted for age, sex, mean FD
```{r calc rew two tp adjusted, eval = FALSE, echo = FALSE}
R1 = lme4::lmer(DMN_FC ~ timepoint*group + age + sex + meanFD (1|subj))
R0 = lme4::lmer(DMN_FC ~ timepoint + group + age + sex + meanFD(1|subj))
```
### reward for three timepoints, adjusted for age, sex, mean FD
```{r calc rew three tp, eval = FALSE, echo = FALSE}
R1 = lmer(DMN_FC ~ timepoint*group + age + sex + meanFD (1|subj))
R0 = lmer(DMN_FC ~ timepoint + group + age + sex + meanFD (1|subj))
```

## Effect of mean and change in BMI (exploratory analysis)

@@

<!-- Neurovault -->
Unthresholded maps for the t-tests as well as contrasts of average BMI and change BMI of the main model, and post-hoc contrasts for average FD and change FD will be are published on NeuroVault.

>>>>

### Aggregated FC values
To extract aggregated FC values, we first calculated the mean DMN and reward networks over all participants and timepoints, adjusted for age and sex. We used global signal regressed data as input and clusterwise bootstrapping with N=999. Network masks were formed from all voxels within clusters which survived a clusterwise multiple comparison correction of p_FWE < 0.05. We extracted the average connectivity from these masks for all data points.

@@

Further, we evaluated the impact of our denoising pipelines. Table \@ref(tab:tableQcfc) shows that in the minimally preprocessed data, mFD and connectivity was significantly positively correlated (mean Spearman's r = `r round(res[1,1],2)`), with a strong distance dependence (mean Spearman's r = `r round(res[1,5],2)`).  Denoising procedures (AROMA, AROMA+CC, AROMA+CC+GSR) gradually reduced the correlation of mFD and connectivity, and also alleviated the distance dependency. AROMA+CC+GSR performed best, with almost no mFD-FC correlation (mean Spearman's r = `r round(res[4,1],2)`), and minimal positive distance dependence (mean Spearman's r = `r round(res[4,5],2)`). Yet, AROMA+CC also performed well, and we thus calculated our analysis models for both preprocessing schemes.
Figure \@ref{fig:figNetworks} shows the DMN and reward network which we used for extracting the aggregated FC valued. Both networks include previously described brain regions (i.e. medial prefrontal cortex, parietal lobules and posterior hippocampus for the DMN, and anterior cingulare, right amygdala and ventromedial-prefrontal cortex for the reward network). Yet, the reward network is less pronounced (overall lower T-values) and symmetric (no significant connectivity with left amygdala), which might be due to SNR differences in the seed brain regions (see Figure \@{ref:fig:figtSNR}.
>>>>>>> include glass brain of networks and LLMMs into draft

```{r figNetworks, echo=FALSE, out.width="49%", out.height="30%", fig.cap="Resting state networks based on AROMA+CC+GSR input data, one-sample t-tests adjusting for age and sex over all timepoints and participants with bootstrapped clusterwise inference (p_FWE<0.05)",fig.subcap=c('DMN (seeded in bilateral precuneus)','Reward network (seeded in bilateral Nucleus Accumbens'),fig.show='hold',fig.align='center'}
networks=c("/data/pt_02161/Results/Project2_resting_state/connectivity/Analysis/aggFC/DMN/DMN_thresholded_for_clusterFWE0.05.png","/data/pt_02161/Results/Project2_resting_state/connectivity/Analysis/aggFC/Rew/Rew_thresholded_for_clusterFWE0.05.png")
knitr::include_graphics(networks)
```
```{r prepare tsnr ROI plot, echo=FALSE}
final$subj.ID_tp=paste0(final$subj.ID,'_', final$tp)

tsnr=read.table("/data/pt_02161/Analysis/Preprocessing/qa/rs_qa/group_level_QA/check_tsnr/tsnr_in_ROIs.txt")
tsnr=tsnr[tsnr$V1!="subj",]
colnames(tsnr)=c("subj.ID_tp","roi","median","mean","sd")
levels(tsnr$roi)=droplevels(tsnr$roi)
levels(tsnr$roi)=c("NAcc", "precuneus")
tsnr$mean=as.double(tsnr$mean)

final_tsnr=merge(final, tsnr, by="subj.ID_tp", all.x=TRUE)

p<-ggplot(aes(as.factor(roi), mean),data=final_tsnr) +
  geom_violin(aes(fill=roi)) + geom_jitter(height = 0, width = 0.1) + xlab("Region of interest") + ylab("average tSNR") + 
  scale_fill_manual(values=wes_palette("Darjeeling2",4,type="discrete")[2:3]) +
  theme(axis.text=element_text(size=10),
  axis.title=element_text(size=12),
  strip.text = element_text(size=12),
  legend.position="")
```

```{r figtSNR,fig.cap="Lower tSNR in the region of interest NAcc versus precuneus used for seed-based connectivity (shown are all 101 timepoints)", fig.width=6, fig.height=4,  echo=FALSE}
p
```

## Interaction effect of intervention and time (preregistered analysis)
### DMN/reward network with whole brain mask (preregistered analysis)

@@

<!-- Neurovault -->
Unthresholded maps for the t-tests of the DMN and reward network (which were used for data aggregation) as well as contrasts of group, time and group-by-time interaction will be published on NeuroVault.

@@

## Interaction effect group by time on aggregated FC (preregistered analysis)
```{r load agg FC data and prepare, echo=FALSE}
agg_FC_DMN=read.table("/data/pt_02161/Results/Project2_resting_state/connectivity/Analysis/aggFC/DMN/agg_FC_DMN.txt")
agg_FC_DMN_ID=read.table("/data/pt_02161/Results/Project2_resting_state/connectivity/Analysis/aggFC/DMN/agg_FC_DMN_ID.txt")
agg_FC_Rew_ID=read.table("/data/pt_02161/Results/Project2_resting_state/connectivity/Analysis/aggFC/Rew/agg_FC_Rew_ID.txt")
agg_FC_Rew=read.table("/data/pt_02161/Results/Project2_resting_state/connectivity/Analysis/aggFC/Rew/agg_FC_Rew.txt")
for (i in c(1:nrow(agg_FC_DMN_ID))){
  agg_FC_Rew[i,"subj.ID_tp"]=strsplit(base::strsplit(as.character(agg_FC_Rew_ID[i,"V1"]),'/')[[1]][10],'_nacc')[[1]][1]
  agg_FC_DMN[i,"subj.ID_tp"]=strsplit(base::strsplit(as.character(agg_FC_DMN_ID[i,"V1"]),'/')[[1]][10],'_pcc')[[1]][1]
}

colnames(agg_FC_Rew)=c("mean_Rew_conn","sd_Rew_conn","subj.ID_tp")
colnames(agg_FC_DMN)=c("mean_DMN_conn","sd_DMN_conn","subj.ID_tp")
final_FC$subj.ID=as.factor(final_FC$subj.ID)

final_FC=merge(final, agg_FC_Rew, by="subj.ID_tp", all.x=TRUE)
final_FC=merge(final_FC, agg_FC_DMN, by="subj.ID_tp", all.x=TRUE)

#calculate mean FD over timepoints
final_FC.meanFD=tapply(X=final_FC$logmFD,
                        INDEX=final_FC$subj.Nr, FUN=mean, na.rm=TRUE)
final_FC$mean.logmFD=
  final_FC.meanFD[as.numeric(final_FC$subj.Nr)]
final_FC$within.logmFD=
  final_FC$logmFD-final_FC$mean.logmFD
```

### DMN for two timepoints
```{r vif for models, echo=FALSE}
test_vif=lm(mean_DMN_conn ~ tp + group_factor + Age_BL + Sex + mean.logmFD, data=final_FC)
vif(test_vif)
```

```{r calc DMN two tp, eval = FALSE, echo = FALSE}
R1 = lme4::lmer(mean_DMN_conn ~ tp*group_factor + (1|subj.ID), data=final_FC[final_FC$tp!="fu2",])
R0 = lme4::lmer(mean_DMN_conn ~ tp + group_factor + (1|subj.ID), data=final_FC[final_FC$tp!="fu2",])
anova(R1,R0)
tab_model(R1,R0)
```

### DMN for two timepoints, adjusted for age, sex, mean FD
```{r calc DMN two tp adjusted, eval = FALSE, echo = FALSE}
R1 = lme4::lmer(mean_DMN_conn ~ tp*group_factor + Age_BL + Sex + logmFD +(1|subj.ID), data=final_FC[final_FC$tp!="fu2",])
R0 = lme4::lmer(mean_DMN_conn ~ tp + group_factor + Age_BL + Sex + logmFD +(1|subj.ID), data=final_FC[final_FC$tp!="fu2",])
anova(R1,R0)
tab_model(R1,R0)
```
### DMN for three timepoints, adjusted for age, sex, mean FD
```{r calc DMN three tp, eval = FALSE, echo = FALSE}
R1 = lme4::lmer(mean_DMN_conn ~ tp*group_factor + Age_BL + Sex + mean.logmFD +(1|subj.ID), data=final_FC)
R0 = lme4::lmer(mean_DMN_conn ~ tp + group_factor + Age_BL + Sex +mean.logmFD +(1|subj.ID), data=final_FC)
anova(R1,R0)
```

```{r DMN conn over three timepoints, echo=FALSE}
fig_DMNconn <-
  ggplot(final_FC, aes(
    x = tp,
    y = mean_DMN_conn,
    group = subj.ID,
    color = condition
  )) +
  geom_line() + geom_point() +
  stat_summary(
    aes(group = condition),
    fun = mean,
    geom  = "line",
    size = 2,
    color = c(rep("#00305E", 3), rep("#C00045", 3))
  ) +
  facet_grid(. ~ condition, labeller = labeller(condition=condition.labs)) + xlab("time points")
fig1=fig_DMNconn +
  scale_x_discrete(labels = c("bl" = "0", "fu" = "6", "fu2" = "12")) +
  scale_color_manual(
    values = c("#00305E77", "#C0004577"),
    labels = c("intervention", "control")
  ) +
  xlab("Time [months]") + ylab ("mean DMN connectivity") +
  theme_bw() +  theme(
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    strip.text = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.position = "none"
  )
fig1
```

### reward
```{r calc rew two tp, eval = FALSE, echo = FALSE}
R1 = lme4::lmer(mean_Rew_conn ~ tp*group_factor + (1|subj.ID), data=final_FC[final_FC$tp!="fu2",])
R0 = lme4::lmer(mean_Rew_conn ~ tp + group_factor + (1|subj.ID), data=final_FC[final_FC$tp!="fu2",])
anova(R1,R0)
```
### reward for two timepoints, adjusted for age, sex, mean FD
```{r calc rew two tp adjusted, eval = FALSE, echo = FALSE}
R1 = lme4::lmer(mean_Rew_conn ~ tp*group_factor + Age_BL + Sex + logmFD +(1|subj.ID), data=final_FC[final_FC$tp!="fu2",])
R0 = lme4::lmer(mean_Rew_conn ~ tp + group_factor + Age_BL + Sex +logmFD +(1|subj.ID), data=final_FC[final_FC$tp!="fu2",])
anova(R1,R0)
```
### reward for three timepoints, adjusted for age, sex, mean FD
```{r calc rew three tp, eval = FALSE, echo = FALSE}
R1 = lme4::lmer(mean_Rew_conn ~ tp*group_factor + Age_BL + Sex + logmFD +(1|subj.ID), data=final_FC)
R0 = lme4::lmer(mean_Rew_conn ~ tp + group_factor + Age_BL + Sex +logmFD +(1|subj.ID), data=final_FC)
anova(R1,R0)
```

```{r Rew conn over three timepoints, echo=FALSE}
fig_Rewconn <-
  ggplot(final_FC, aes(
    x = tp,
    y = mean_Rew_conn,
    group = subj.ID,
    color = condition
  )) +
  geom_line() + geom_point() +
  stat_summary(
    aes(group = condition),
    fun = mean,
    geom  = "line",
    size = 2,
    color = c(rep("#00305E", 3), rep("#C00045", 3))
  ) +
  facet_grid(. ~ condition, labeller = labeller(condition=condition.labs)) + xlab("time points")
fig2=fig_Rewconn +
  scale_x_discrete(labels = c("bl" = "0", "fu" = "6", "fu2" = "12")) +
  scale_color_manual(
    values = c("#00305E77", "#C0004577"),
    labels = c("intervention", "control")
  ) +
  xlab("Time [months]") + ylab ("mean reward network connectivity") +
  theme_bw() +  theme(
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    strip.text = element_text(size = 12),
    legend.text = element_text(size = 10),
    legend.position = "none"
  )
fig2
```
## Effect of mean and change in BMI (exploratory analysis)

@@
